\mytitle

### 2012-02-22 papers/SAS2012/implicitization.tex
Encore du compactage

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph}).
\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries\label{contr:disjunctive} (See \S\ref{sec:disjunctive}).
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
It uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which
supports a variety of abstract domains for numerical variables, from
which we can choose with minimal changes to our analyzer.

@@

\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
\end{enumerate}

@@

Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
\footnote{%
Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\begin{equation}
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\label{eqn:transformer_correct}
\end{equation}
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
associated to $p_i$ :
\[
f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j^s)
\]
The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
Also, to prevent the algorithm from adding too many paths at a time,
and in particular to prevent the algorithm from trying to re-add the
same path infinitely many times, we use another abstract value
associated to the control point $p_j$, noted $X_j^d$, which is distinct
from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
is the starting point of the path, and $X_j^d$ when it is its destination point.
We thus check the satisfiability of the formula
$f'(p_i)$, where:
\[
f'(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
\neg X_j^d)
\]
$X_j^d$ is updated when the point $p_j$ is the target of a new path.
This way, further SMT queries do not
compute other paths with the same source and destination if it is not
needed (because these new paths would not make $X_j^d$ grow, hence
would not be returned by the SMT solver).

@@

\begin{enumerate}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_{10}$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7, 8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_{10}$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{enumerate}

@@

Our method improves this technique in two ways :
\begin{itemize}
\item Instead of enumerating the whole set of paths, we keep them implicit and
compute them only when needed.

\item At each loop iteration of the original algorithm \cite{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
In our approach, we compute such an image only if it makes the resulting abstract value grow.
\end{itemize}

@@

The $\sigma_i$ functions take as parameters the index of the starting abstract
value, and the path we focus on. 
Since we dynamically construct these functions
during the analysis, we store their already computed image into a
compact representation, such as Algebraic Decision Diagrams.
$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
only once.
When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
the image of the
abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
If such an index exists, then we set $\sigma_i(j,k) = j'$.
Otherwise:
\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{itemize}
The main difference with the original
algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
exponential number of paths to find a solution.

>>>>

\renewcommand{\baselinestretch}{1}

@@

This article makes the following contributions:
\begin{compactenum}
\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph}).
\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries\label{contr:disjunctive} (See \S\ref{sec:disjunctive}).
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
It uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which
supports a variety of abstract domains for numerical variables, from
which we can choose with minimal changes to our analyzer.

@@

\item We conducted extensive experiments with this tool, on real-life programs\avirer{: (i) fixing the abstract domain, varying the iteration technique (\S\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains})}.
\end{compactenum}

@@

Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$. The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}. An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion. \avirer{In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.}
The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
%
% \footnote{%
% Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents. This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}

@@

Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g. $x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
\(
\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
x' \in \gamma \circ \abstr{\tau}(\abstr{x})
\).
It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.

@@

If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
associated to $p_i$ :
\vspace{-.5em}
\[
f(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
 \neg X_j^s)
\vspace{-.1em}
\]
The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
Also, to prevent the algorithm from adding too many paths at a time,
and in particular to prevent the algorithm from trying to re-add the
same path infinitely many times, we use another abstract value
associated to the control point $p_j$, noted $X_j^d$, which is distinct
from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
is the starting point of the path, and $X_j^d$ when it is its destination point.
We thus check the satisfiability of the formula
$f'(p_i)$, where:
\vspace{-.5em}
\[
f'(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
\neg X_j^d)
\vspace{-1em}
\]
$X_j^d$ is updated when the point $p_j$ is the target of a new path.
This way, further SMT queries do not
compute other paths with the same source and destination if it is not
needed (because these new paths would not make $X_j^d$ grow, hence
would not be returned by the SMT solver).

@@

\begin{compactenum}
	\item The first subset of paths is depicted on Figure
		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
	\item
		We then compute the set $P'$ of paths that have to be added into the
		next subgraph. 
		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
		itself, and that goes through the \emph{else} branch of each 
		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
		is then added to our subgraph. Moreover, there is no other path whose
		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
		Additionally, the path that goes from $p_5$ to
		$p_{10}$, that goes through the \emph{else} branch of each 
		\emph{if-then-else} at lines 7, 8 and 9, is also 
		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
		image outside $-10 \leq \lstinline|x_old| \leq 10$.
	\item
		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
		Step 2). When we have reached an inductive invariant, we narrow, and the
		result for both $p_5$ and $p_{10}$ is the polyhedron 
		$-10000 \leq \lstinline|x_old| \leq 10000$.
	\item Finally, we compute the new subgraph. The SMT-solver does not find any
		new path that makes the abstract values grow, and the algorithm
		terminates.
\end{compactenum}

@@

Our method improves this technique in two ways :
\begin{compactitem}
\item Instead of enumerating the whole set of paths, we keep them implicit and
compute them only when needed.

\item At each loop iteration of the original algorithm \cite{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
In our approach, we compute such an image only if it makes the resulting abstract value grow.
\end{compactitem}

@@

The $\sigma_i$ functions take as parameters the index of the starting abstract
value, and the path we focus on. 
Since we dynamically construct these functions
during the analysis, we store their already computed image into a
compact representation, such as Algebraic Decision Diagrams.
$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
only once.
When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
the image of the
abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
If such an index exists, then we set $\sigma_i(j,k) = j'$.
Otherwise:
\begin{compactitem}
	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
\end{compactitem}
The main difference with the original
algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
exponential number of paths to find a solution.
